# Hierarchical Spatial Data

Non-hierarchical spatial data are points or vertices in one dimension and edges
or paths in two dimensions. They are the only spatial entities which may be
considered atomic; that is, not composed of anything else. Two vertices may form
an path, or they may simply be two unrelated vertices (spatial autocorrelation
notwithstanding). Whether they form a path or not will often depend on the
intention of the person considering those vertices  Thus we quickly arrive at
the conclusion that spatial hierarchies are in the eye of the beholder. If the
object or objective of an analysis concerns relationships between vertices, then
the path connecting two vertices is likely to be important.  Storing information
on the system comprised of those two vertices then requires storing information
on the vertices themselves, as well as on the way they are connected or related;
that is, on the path 

The `sphier` project aims to provide a common form for representing hierarchical
data of any form, but is particularly oriented toward spatial data. The primary
`sphier` entities are `vertex`, `path`, `object`, and `container`. The first
three are equivalent to `silicate` `vertex`, `path`, and `object` entities,
while the latter is analogous to, yet somewhat different from,
a `silicate::path_link_vertex` table.  Note that there are no `sphier`
`coordinate` objects, as coordinates are contained within the `vertex` objects,
along with a primary index enumerating the vertices. Silicate references
vertices by position only, yet the necessity of referencing by index in
`sphier` will be demonstrated below.  The following concrete example illustrates
this `sphier` structure:
```{r ex1-sphier, eval = FALSE}
p1 <- c (0, 01) # x, y coordinates
p2 <- c (1, 1)
s <- sphier (rbind (p1, p2), type = "a kind of line")
s
```
```{r ex1-fake, echo = FALSE}
vertex <- tibble::tibble (
        index = c (1, 2),
        x = c (0, 1),
        y = c (0, 1))
path <- tibble::tibble (
        path = c (1L, 1L),
        vertex = c (1L, 2L))
container <- tibble::tibble (
        object = 1L,
        type = "path",
        contains = NA)
object <- tibble::tibble (
        object = 1L,
        type = "a kind of line",
        value = sqrt (2))
list (vertex = vertex, path = path, container = container, object = object)
```
The first column of the `container` object enumerates the number of the
container, and is directly used to construct higher-order hierarchical objects
as illustrated below.  The first column of `object` indexes directly into the
equivalent `object` column of `container`.  In this case, there is only one
single container, numbered 1, and that object describing that container holds
the distance between the two vertices. Note that the `type` of object in
`container` is `path`, not `a kind of line`. Containers use the two keywords
`path` and `vertex` to refer to objects of those respective types, with these
types - and only these types - being empty containers.

# An Hierarchical Example

The "a kind of line" object in the preceding example was not hierarchical and
therefore did not "contain" anything. Consider the addition of another
(non-hierarchical) object giving the following `sphier` object:
```{r ex2-sphier, eval = FALSE}
p1 <- c (2, 3)
p2 <- c (2, 3)
s <- c (s, sphier (rbind (p1, p2), type = "a kind of line"))
s
```
```{r ex2-fake, echo = FALSE}
vertex <- tibble::tibble (
        indx = 1:4,
        x = 0:3,
        y = 0:3)
path <- tibble::tibble (
        path = c (1L, 1L, 2L, 2L),
        vertex = c (1L, 2L, 3L, 4L))
container <- tibble::tibble (
        object = 1:2,
        type = rep ("path", 2),
        contains = rep (NA, 2))
object <- tibble::tibble (
        object = 1:2,
        type = rep ("a kind of line", 2),
        value = rep (sqrt (2)), 2)
list (vertex = vertex, path = path, container = container, object = object)
```
Those two separate lines may be considered to form part of some higher-level "metaline" object, represented as:
```{r ex3-sphier, eval = FALSE}
s <- c (s, sphier (object = c (1, 2), type = "metaline",
                   relationship = "thick"))
```
```{r ex3-fake, echo = FALSE}
vertex <- tibble::tibble (
        indx = 1:4,
        x = 0:3,
        y = 0:3)
path <- tibble::tibble (
        path = c (1L, 1L, 2L, 2L),
        vertex = c (1L, 2L, 3L, 4L))
container <- tibble::tibble (
        object = c (1, 2, 3, 3),
        type = c (rep ("path", 2), rep ("metaline", 2)),
        contains = c (NA, NA, 1, 2))
object <- tibble::tibble (
        number = 1:3,
        type = c (rep ("a kind of line", 2), "metaline"),
        value = c (rep (sqrt (2), 2), NA),
        relationships = c (NA, NA, "thick"))
list (vertex = vertex, path = path, container = container, object = object)
```
In this case, the "metaline" object contains the two basic paths of two
vertices.  each, and declares that they share a "thick" relationship.  The
`vertex` object is the only aspect that geometrically anchors the entire data
structure. The `container` object maps inter-relationships between all objects
(both `path` and higher-level objects), and `object` includes the properties
whether quantitative or qualitative of all higher-level objects.

Although `vertex` entities are considered inherently primitive objects that do
not possess attributes, attributes of vertices may be readily stored by
creating associated objects, as in the following example.
```{r ex4-sphier, eval = FALSE}
s <- c (s, sphier (vertex = 1, type = "vertex", colour = "blue"))
```
```{r ex4-fake, echo = FALSE}
vertex <- tibble::tibble (
        indx = 1:4,
        x = 0:3,
        y = 0:3)
path <- tibble::tibble (
        path = c (1L, 1L, 2L, 2L),
        vertex = c (1L, 2L, 3L, 4L))
container <- tibble::tibble (
        object = c (1:3, 3:4),
        type = c (rep ("path", 2), rep ("metaline", 2), "point"),
        contains = c (NA, NA, 1, 2, NA))
object <- tibble::tibble (
        number = 1:4,
        type = c (rep ("a kind of line", 2), "metaline", "point"),
        value = c (rep (sqrt (2), 2), NA, NA),
        relationships = c (NA, NA, "thick", NA),
        colour = c (rep (NA, 3), "blue"))
list (vertex = vertex, path = path, container = container, object = object)
```
This example also illustrates that the more information that is stored regarding
the various `object` entities in a given spatial representation, the less
efficient the object storage may ultimately become. Nevertheless, the storage of
all data in this scheme of four flat tables is extremely advantageous for
executing geometric operations on objects, paths, or vertices, because these
operations can be directly translated into filtering and joining operations
applied directly to these tables.


# Time; trajectories; spatial dynamics

The `sphier` system explicitly accommodates representations of temporal
dynamics. As the simplest level, a vector trajectory can be represented through
extension of the `vertex` component to include a `t` column. The maximum that
the `sphier` `vertex` table can contain is the four columns of (`x`, `y`, `z`,
`t`).

